Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]

### 2020-05-15 Covid_ethnicity_Scotland.Rmd
fixed mislabelling in ethnicity report

# Methods 
All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems. This system records CHI but not ethnicity.

@@

## Coding of ethnicity
Names of cases and controls were available to the CHI team. They used the ONOMAP software to assign predicted ethnicity to all cases and controls. This was output in anonymised format to the Modelling & Research Cell linkage team. ONOMAP is a software package developed by Mateos at UCL which used people's first and family names to estimate underlying ethnic group. It has been developed using algorithms built from the Worldnames database that covered both names and ethnic group in a large number of countries. ONOMAP assigns both predicted geographical region of origin and ethnicity. ONOMAP has previously been reported to have variables sensitivity for assigning non white ethnic groups 17-89%. ([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were allocated to ethnic groups. Names classified "Muslim" in the ONOMAP classification were assigned to the South Asian group, as Scotland most people with Muslim names are South Asian.  The ONOMAP classification assigns a high proportion of names in Scotland to the category "Celtic" which does not distinguish Irish from Scottish names so that White Irish cannot be differentiated from White British.

@@

A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded (Clark & Buchanan ISD 2017).  The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.

@@

## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population

rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
knitr::kable(table.severe.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

As shown above the numbers of persons testing positive from non-White ethnic groups is low. There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites. The frequency of severe disease was lower in South Asans but with these small numbers this estimate is unstable.

## Relation of hospitalised COVID-19 to ethnicity in the Scottish population

rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
knitr::kable(table.hospitalized.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

@@

# Supplementary tables
These tables including test-positive individuals who were not hospitalized are included by request.  They should be interpreted with caution, as selection for testing is influenced by many factors including health care worker status.

## Relation of severity of COVID-19 to ethnicity among test-positive cases

@@

pander(testpositives.ethnic, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")

pander(testpositives.ethnic.smr, missing=".",
       split.tables=Inf,
       caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")

@@

pander(testpositives.carehome, missing=".",
       split.tables=Inf,
       caption="Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")

@@

pander(testpositives.healthboard, missing=".",
       split.tables=Inf,
       caption="Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")

@@

## Relation of test-positive status to ethnicity in the Scottish population

rownames(table.testpositives.demog) <- gsub("_", " ", rownames(table.testpositives.demog))
knitr::kable(table.testpositives.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.onomap",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.testpositives.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.ethnicsmr",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

>>>>

# Methods 
The REACT-SCOT case-control study was set up to investigate risk factors for severe COVID-19 and to lay the basis for developing risk stratification algorithms. All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems. This system records CHI but not ethnicity.

@@

## Coding of ethnicity
Names of cases and controls were available to the CHI team. They used the ONOMAP software to assign predicted ethnicity to all cases and controls. This was output in anonymised format to the Modelling & Research Cell linkage team. ONOMAP is a software package developed by Mateos at UCL which used people's first and family names to estimate underlying ethnic group. It has been developed using algorithms built from the Worldnames database that covered both names and ethnic group in a large number of countries. ONOMAP assigns both predicted geographical region of origin and ethnicity. ONOMAP has previously been reported to have variable sensitivity for assigning non-White ethnic groups. ([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were used to assign ethnicity. Names classified "Muslim" in the ONOMAP classification were assigned to the South Asian group, as Scotland most people with Muslim names are South Asian.  The ONOMAP classification assigns a high proportion of names in Scotland to the category "Celtic" which does not distinguish Irish from Scottish names so that White Irish cannot be differentiated from White British.

@@

A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded (Clark & Buchanan ISD 2017).  The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.  The Black category combines those who self-identify as Black British, Black Caribbean, Black African and Black Other.

The rate ratios shown are based on conditional logistic regression.  With the incidence density sampling used in this study, the odds ratios in a conditional logistic regression are equal to rate ratios.  The unconditional odds ratios calculated from the numbers of cases and controls in the tables are not the same as the conditional odds ratios.

@@

## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population
Tables \ref{tab:testpos.ethnicsmr}. \ref{tab:ethnicsmr.hosp} and \ref{tab:ethnicsmr} compare cases and controls with ethnicity based on SMR classificaiton for all test positive cases, all hospitalized cases, and severe cases.

Table \ref{tab:testpos.onomap}, \ref{tab:onomap.hosp}, \ref{tab:onomap} show the same comparisons with ethnicity based on ONOMAP name classification

There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites. The rate of severe disease in South Asians was about half the rate in Whites.

Tables \ref{tab:severity.smr} and \ref{tab:severity} show the distribution over severity categories by ethnicity

Tables \ref{tab:carehome} and \ref{tab:board} show the distribution of care home status and health board.

@@

# Tables
Tables comparing all test-positive individuals with controls are included by request.  They should be interpreted with caution, as selection for testing is influenced by many factors including health care worker status.  Similar considerations apply to comparisons of disease severity among those who test positive.

rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
knitr::kable(table.testpositives.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.ethnicsmr",
             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm")  %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")
```

rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
knitr::kable(table.hospitalized.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```{r ethnic5smr, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic5smr) <- gsub("_", " ", rownames(table.ethnic5smr))
knitr::kable(table.ethnic5smr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnic5smr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

pander(testpositives.ethnic.smr, missing=".",
       split.tables=Inf,
       caption="\\label{tab:severity.smr}Severity of disease in test-positive cases, by ethnicity assigned in SMR records")

rownames(table.testpositives.demog) <- gsub("_", " ", rownames(table.testpositives.demog))
knitr::kable(table.testpositives.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.onomap",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.testpositives.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity based on name classification") %>%
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    column_spec(column=2, width="1.95cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    column_spec(column=2, width="1.95cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
knitr::kable(table.severe.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

pander(testpositives.ethnic, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="\\label{tab:severity}Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")

@@

pander(testpositives.carehome, missing=".",
       split.tables=Inf,
       caption="\\label{tab:carehome}Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")

@@

pander(testpositives.healthboard, missing=".",
       split.tables=Inf,
       caption="\\label{tab:board}Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")

@@

### 2020-05-15 casecontrol.Rmd
fixed mislabelling in ethnicity report

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(10*exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

>>>>

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.
